Tropical Andes Taxonomic Diversity of Plants and Frugivores

### 2024-05-14 code/L2/L2_1_TropicalAndes_taxdiv.Rmd
move code to L1_LastStep_Diversity_Input_Objects script

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(smoothr)
library(purrr)
library(raster)
library(scico)
library(ggspatial)
```

@@

# Polygons of countries
```{r}
worldMap <- ne_countries(scale = "medium", type = "countries", returnclass = "sf")
Americas <- ne_countries(continent = c("North America", "South America"), returnclass = "sf")
#polygon of Tropical Andes
TApoly <- worldMap %>% filter(sovereignt == "Bolivia" |sovereignt == "Ecuador" | sovereignt == "Venezuela" | sovereignt == "Colombia" | sovereignt == "Peru")
```

# Convert data to spatial data
```{r}
plants.sf <- st_as_sf(TropicalAndes_plant_occ_forest, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

frugivores.sf <- st_as_sf(TropicalAndes_frugivore_occ_forest, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
```

# Transform to projected coordinate reference system (units from degrees to meters)
```{r}
Americas <- st_transform(Americas, 5389)
TApoly <- st_transform(TApoly, 5389)
TropicalAndes_IUCNHabitat_Forest <- st_transform(TropicalAndes_IUCNHabitat_Forest, 5389)
plants.sf <- st_transform(plants.sf, 5389)
frugivores.sf <- st_transform(frugivores.sf, 5389)
```

# Group by species
```{r}
plants_sf_species <- plants.sf %>%
  group_by(species) %>%
  summarise()

frugivores_sf_species <- frugivores.sf %>%
  group_by(species) %>%
  summarise()
```

# Plot Base Map
```{r}
basePlot <-
ggplot() +
  geom_sf(data = Americas, fill = "white") +
  geom_sf(data = TApoly) +
  geom_sf(data = TropicalAndes_IUCNHabitat_Forest, fill = "forestgreen")+
  labs(title = "Tropical Andes", x = "Latitude", y = "Longitude") +
  coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
    pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
  theme(panel.background = element_rect(fill = "lightblue"))

basePlot
```

# Plot points
```{r}
plantsPointsPlot <-
ggplot() +
  geom_sf(data = Americas, fill = "white") +
  geom_sf(data = TApoly) +
  geom_sf(data = plants_sf_species, pch = 21) +
  labs(title = "Tropical Andes Plant Occurrences", x = "Latitude", y = "Longitude") +
  coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
    pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
  theme(panel.background = element_rect(fill = "lightblue"))

plantsPointsPlot
```

```{r}
frugivoresPointsPlot <-
ggplot() +
  geom_sf(data = Americas, fill = "white") +
  geom_sf(data = TApoly) +
  geom_sf(data = frugivores_sf_species, pch = 21) +
  labs(title = "Tropical Andes Frugivore Occurrences", x = "Latitude", y = "Longitude") +
  coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
    pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
  theme(panel.background = element_rect(fill = "lightblue"))

frugivoresPointsPlot
```

```{r}
#check units
st_crs(TApoly, parameters = TRUE)$units_gdal
st_crs(TApoly)
st_crs(TropicalAndes_IUCNHabitat_Forest)
```

# Plot TD/Richness
function for richness plots
```{r}
create_rich_plots <- function(resolution_meters) {
 # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

@@

  # Helper function to generate plots
  generate_plot <- function(data, title) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_scico(palette = "romaO", direction = 1) +
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5) +
      annotation_north_arrow(location = "bl", which_north = "true", 
                             pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  # Generate plots
  plant_plot <- generate_plot(plant_richness_grid, paste("Tropical Andes Plant Richness [", resolution_meters/1000, " km]", sep=""))
  frugivore_plot <- generate_plot(frugivore_richness_grid, paste("Tropical Andes Frugivore Richness [", resolution_meters/1000, " km]", sep=""))

  # Compare plant and frugivore richness
  compare_richness_summary <- summary(
    lm(frugivore_richness_grid$num_species ~ plant_richness_grid$num_species))

@@

  # Create scatter plot
  plant_frug_richness_compare <- ggplot(comparison_data,
                                        aes(x = Plant_Richness, y = Frugivore_Richness)) +
    geom_point() +
    labs(title = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]",
                       sep = ""), x = "Plant Richness by cell", y = "Frugivore Richness by cell") +
    geom_smooth(method = "lm", se = FALSE, color = "red")

@@

  # Return a list of plots
  list(plant_plot = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid, 
       compare_richness_summary = compare_richness_summary,
       plant_frug_richness_compare = plant_frug_richness_compare,
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}

@@

## 1 km resolution
```{r}
richness_1km <- create_rich_plots(1000)
# save plots
plantgridRichTA_1km <- plant_plot
frugivoregridRichTA_1km <- frugivoregridRichTA
# save richness data
plant_cellRichness_1km <- plant_cellRichness
frugivore_cellRichness_1km <- frugivore_cellRichness

compare_richness_r2_1km <- compare_richness_summary$r.squared
```

## 5 km resolution
```{r}
richness_5km <- create_rich_plots(5000)
# save plots
plantgridRichTA_5km <- plantgridRichTA
frugivoregridRichTA_5km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_5km <- plant_cellRichness
frugivore_cellRichness_5km <- frugivore_cellRichness

compare_richness_r2_5km <- compare_richness_r2
```

## 10 km
```{r}
richness_10km <- create_rich_plots(10000)
# save plots
plantgridRichTA_10km <- plantgridRichTA
frugivoregridRichTA_10km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_10km <- plant_cellRichness
frugivore_cellRichness_10km <- frugivore_cellRichness

compare_richness_r2_10km <- compare_richness_r2
```

## 25 km
```{r}
richness_25km <- create_rich_plots(25000)
# save plots
plantgridRichTA_25km <- plantgridRichTA
frugivoregridRichTA_25km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_25km <- plant_cellRichness
frugivore_cellRichness_25km <- frugivore_cellRichness

compare_richness_r2_25km <- compare_richness_r2
```

## 50 km
```{r}
richness_50km <- create_rich_plots(50000)
# save plots
plantgridRichTA_50km <- plantgridRichTA
frugivoregridRichTA_50km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_50km <- plant_cellRichness
frugivore_cellRichness_50km <- frugivore_cellRichness

compare_richness_r2_50km <- compare_richness_r2
```

## 75 km 
```{r}
richness_75km <- create_rich_plots(75000)
# save plots
plantgridRichTA_75km <- plantgridRichTA
frugivoregridRichTA_75km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_75km <- plant_cellRichness
frugivore_cellRichness_75km <- frugivore_cellRichness

compare_richness_r2_75km <- compare_richness_r2
```

## 100 km 
```{r}
richness_100km <- create_rich_plots(100000)
# save plots
plantgridRichTA_100km <- plantgridRichTA
frugivoregridRichTA_100km <- frugivoregridRichTA
## Extract cell values
plant_cellRichness_100km <- plant_cellRichness
frugivore_cellRichness_100km <- frugivore_cellRichness

compare_richness_r2_100km <- compare_richness_r2
```

# Compare r-squared by resolution
Plot
```{r}
# create dataframe of resolutions and r-squared values
resolution <- c("5", "25", "50", "75", "100")
r_squared <- c(compare_richness_r2_5km, compare_richness_r2_25km, compare_richness_r2_50km, compare_richness_r2_75km, compare_richness_r2_100km)
r_squared_compare <- data.frame(resolution, r-squared)
```

```{r}
ggplot(r_squared_compare, aes(x=resolution, y=r_squared)) +
  geom_point()+
  labs(title = "R-squared values of plant-frugivore richness relationships by resolution", x = "Resolution (km)", y = "R-squared",
       fill = "Number of species") 
```

# Multi panel plots
Richness maps
```{r}
grid.arrange(plot_1, plot_2, nrow = 1)
```

# Write data to csv
```{r}
write.csv(plant_cellRichness0.75, file.path(output_path,"TropicalAndes_GBIF_plant_richness_sub.csv"))
write.csv(frugivore_cellRichness0.75, file.path(output_path,"TropicalAndes_frugivore_richness_sub.csv"))
```

>>>>

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(smoothr)
library(purrr)
library(raster)
library(ggspatial)
library(gridExtra)
```

@@

# Plot TD/Richness
## Function for richness plots
Need to set legend scale and color for plants constant (same for frugivores)
remove cells with richness of 0
```{r}
create_rich_plots <- function(resolution_meters) {
 # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

@@

  # Helper function to generate plots
  generate_plot <- function(data, title) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_scico(palette = "romaO", direction = 1) +
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-85, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  # Generate plots
  plant_plot <- generate_plot(plant_richness_grid, paste("Tropical Andes Plant Richness [", resolution_meters/1000, " km]", sep=""))
  plant_plot <- generate_plot(frugivore_richness_grid, paste("Tropical Andes Frugivore Richness [", resolution_meters/1000, " km]", sep=""))

  # Extract cell values
  plant_cellRichness <- plant_plot[["data"]]
  frugivore_cellRichness <- plant_plot[["data"]]

  # Compare plant and frugivore richness
  compare_richness_summary <- summary(
    lm(frugivore_richness_grid$num_species ~ plant_richness_grid$num_species))
  r_squared <- compare_richness_summary$r.squared

@@

  ## Create scatter plot
  plant_frug_richness_compare <- ggplot(comparison_data,
                                        aes(x = Plant_Richness, y = Frugivore_Richness)) +
  geom_point() +
  labs(title = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]", sep = ""),
       subtitle =  paste("R-squared = ", round(r_squared, 2), sep = ""),
       x = "Plant Richness by cell", y = "Frugivore Richness by cell") +
  geom_smooth(method = "lm", se = FALSE, color = "red")  # Add regression line

@@

  # Return a list of plots
  list(plantgridRichTA = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid, 
       compare_richness_summary = compare_richness_summary,
       plant_frug_richness_compare = plant_frug_richness_compare,
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}

@@

## 1 km resolution
```{r}
richness_1km <- create_rich_plots(1000)
print(richness_1km)

# save plots
plantgridRichTA_1km <- richness_1km$plant_plot
frugivoregridRichTA_1km <- richness_1km$frugivoregridRichTA
# save richness data
plant_cellRichness_1km <- richness_1km$plant_cellRichness
frugivore_cellRichness_1km <- richness_1km$frugivore_cellRichness

compare_richness_r2_1km <- richness_1km$compare_richness_summary$r.squared
```

## 5 km resolution
```{r}
richness_5km <- create_rich_plots(5000)
print(richness_5km)

# save plots
plantgridRichTA_5km <- richness_5km$plantgridRichTA
frugivoregridRichTA_5km <- richness_5km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_5km <- richness_5km$plant_cellRichness
frugivore_cellRichness_5km <- richness_5km$frugivore_cellRichness

compare_richness_r2_5km <- richness_5km$compare_richness_summary$r.squared
```

## 10 km
```{r}
richness_10km <- create_rich_plots(10000)
print(richness_10km)

# save plots
plantgridRichTA_10km <- richness_10km$plantgridRichTA
frugivoregridRichTA_10km <- richness_10km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_10km <- richness_10km$plant_cellRichness
frugivore_cellRichness_10km <- richness_10km$frugivore_cellRichness

compare_richness_r2_10km <- richness_10km$compare_richness_summary$r.squared
```

## 25 km
```{r}
richness_25km <- create_rich_plots(25000)
print(richness_25km)

# save plots
plantgridRichTA_25km <- richness_25km$plantgridRichTA
frugivoregridRichTA_25km <- richness_25km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_25km <- richness_25km$plant_cellRichness
frugivore_cellRichness_25km <- richness_25km$frugivore_cellRichness

compare_richness_r2_25km <- richness_25km$compare_richness_summary$r.squared
```

## 50 km
```{r}
richness_50km <- create_rich_plots(50000)
print(richness_50km)
# save plots
plantgridRichTA_50km <- richness_50km$plantgridRichTA
frugivoregridRichTA_50km <- richness_50km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_50km <- richness_50km$plant_cellRichness
frugivore_cellRichness_50km <- richness_50km$frugivore_cellRichness

compare_richness_r2_50km <- richness_50km$compare_richness_summary$r.squared
```

## 75 km 
```{r}
richness_75km <- create_rich_plots(75000)
print(richness_75km)
# save plots
plantgridRichTA_75km <- richness_75km$plantgridRichTA
frugivoregridRichTA_75km <- richness_75km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_75km <- richness_75km$plant_cellRichness
frugivore_cellRichness_75km <- richness_75km$frugivore_cellRichness

compare_richness_r2_75km <- richness_75km$compare_richness_summary$r.squared
```

## 100 km 
```{r}
richness_100km <- create_rich_plots(100000)
print(richness_100km)
# save plots
plantgridRichTA_100km <- richness_100km$plantgridRichTA
frugivoregridRichTA_100km <- richness_100km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_100km <- richness_100km$plant_cellRichness
frugivore_cellRichness_100km <- richness_100km$frugivore_cellRichness

compare_richness_r2_100km <- richness_100km$compare_richness_summary$r.squared
```

# Multi panel plots
Richness maps
```{r}
all_richness_plots <- gridExtra::grid.arrange(#plantgridRichTA_1km, frugivoregridRichTA_1km,
                                              #plantgridRichTA_5km, frugivoregridRichTA_5km, 
                                               plantgridRichTA_10km, frugivoregridRichTA_10km,
                                               plantgridRichTA_25km, frugivoregridRichTA_25km,
                                               plantgridRichTA_50km, frugivoregridRichTA_50km,
                                               plantgridRichTA_75km,  frugivoregridRichTA_75km,
                                               plantgridRichTA_100km, frugivoregridRichTA_100km,
                                               ncol = 5, nrow = 2)
all_richness_plots
ggsave("all_richness_plots.pdf", all_richness_plots, path = output_path, width = 20, height = 8.5, units = "in")
```

# Compare r-squared by resolution
Plot
```{r}
# create dataframe of resolutions and r-squared values 
#[ADD 1, 5,]
resolution <- c( 10, 25, 50, 75, 100)
#ADD compare_richness_r2_1km, compare_richness_r2_5km,
r_squared <- c( compare_richness_r2_10km, compare_richness_r2_25km, compare_richness_r2_50km, compare_richness_r2_75km, compare_richness_r2_100km)
r_squared_compare <- data.frame(resolution, r_squared)

compare_r2_by_res <- ggplot(r_squared_compare, aes(x=resolution, y=r_squared)) +
    geom_point()+
    labs(title = "R-squared values of plant-frugivore richness relationships by resolution",
         x = "Resolution (km)", y = "R-squared", fill = "Number of species")
print(compare_r2_by_res)
```

# Write data to csv
```{r}
write.csv(plant_cellRichness_1km, file.path(output_path,"TropicalAndes_plantRichness_1km.csv"))
write.csv(frugivore_cellRichness_1km, file.path(output_path,"TropicalAndes_frugivoreRichness_1km.csv"))
```

```{r}
write.csv(plant_cellRichness_5km, file.path(output_path,"TropicalAndes_plantRichness_5km.csv"))
write.csv(frugivore_cellRichness_5km, file.path(output_path,"TropicalAndes_frugivoreRichness_5km.csv"))
```

```{r}
write.csv(plant_cellRichness_10km, file.path(output_path,"TropicalAndes_plantRichness_10km.csv"))
write.csv(frugivore_cellRichness_10km, file.path(output_path,"TropicalAndes_frugivoreRichness_10km.csv"))
```

```{r}
write.csv(plant_cellRichness_25km, file.path(output_path,"TropicalAndes_plantRichness_25km.csv"))
write.csv(frugivore_cellRichness_25km, file.path(output_path,"TropicalAndes_frugivoreRichness_25km.csv"))
```

```{r}
write.csv(plant_cellRichness_50km, file.path(output_path,"TropicalAndes_plantRichness_50km.csv"))
write.csv(frugivore_cellRichness_50km, file.path(output_path,"TropicalAndes_frugivoreRichness_50km.csv"))
```

```{r}
write.csv(plant_cellRichness_75km, file.path(output_path,"TropicalAndes_plantRichness_75km.csv"))
write.csv(frugivore_cellRichness_75km, file.path(output_path,"TropicalAndes_frugivoreRichness_75km.csv"))
```

```{r}
write.csv(plant_cellRichness_100km, file.path(output_path,"TropicalAndes_plantRichness_100km.csv"))
write.csv(frugivore_cellRichness_100km, file.path(output_path,"TropicalAndes_frugivoreRichness_100km.csv"))
```
